Modeling the Linux page cache for accurate simulation of data-intensive applications

### 2020-12-11 paper_io_sim.tex

        \subsection{Simulators}

@@

    The Memory Manager simulates two parallel threads: the main one
    implements flushing, eviction, and cached I/Os synchronously, whereas
    the second one, operating in the background, periodically searches for
    expired dirty data in page cache LRU lists and flushes them to disk. We
    use existing storage simulation models~\cite{lebre2015} to simulate disk and
    memory, characterized by their storage capacity, read and write
    bandwidths, and latency. Reusing existing storage models allows us to
    simulate bandwidth sharing between concurrent memory accesses and concurrent disk accesses.

@@

    In the Linux kernel, page cache LRU lists contain file pages. However,
    due to the large number of file pages, simulating lists of pages
    induces substantial overheads.
    Therefore, we introduce the concept of a data block as a unit to represent data
    cached in memory. A data block is a subset of file pages stored in
    page cache that were accessed in the same I/O operation.
    A data block has information about the file name, block size, last access
    time, a dirty flag that represents whether the data is clean (0)
    or dirty (1), and an entry (creation) time.
    Blocks can have different sizes and a given file can have multiple
    data blocks in page cache. In addition, a data block can be split into an
    arbitrary number of smaller blocks.
    \begin{figure}
           \centering
           \includegraphics[width=\columnwidth]{figures/lru_lists.pdf}
           \caption{Model of page cache LRU lists with data blocks.}
           \label{fig:lrulist}
    \end{figure}

@@

    Our simulation model supports chunk-by-chunk file accesses
    with a user-defined chunk size. However, for simplicity, we assume that file pages are
    accessed in a round-robin fashion rather than fully randomly.
    Therefore, when a file is read, cached data is read only after all uncached data was read, and data from the inactive list is read
    before data from the active list
    (data reads occur from left to right in Figure~\ref{fig:read_order}).
    When a chunk of \emph{uncached} data is read, a new clean block is created
    and appended to the inactive list.
    When a chunk of \emph{cached} data is read, one or more existing data blocks in the LRU lists are accessed.
    If these blocks are clean, we merge them together, update the access time and size of the resulting block,
    and append it to the active list.
    If the blocks are dirty, we move them independently to the active list, to preserve their entry time.
    Because the chunk and block sizes may be different, there are situations
    where a block is not entirely read.
    In this case, the block is split in two smaller blocks and one of them is re-accessed.
    % From Tristan: not sure where to put this nor if it's necessary:
    % , in which chunks are read/written
    % until file is entirely read/written.
    \begin{figure}
           \centering
           \includegraphics[width=\columnwidth]{figures/read_order.pdf}
           \caption{File data read order. Data is read from left to right: uncached data
           is read first, followed by data from the inactive list, and finally data from the active list. }
           \label{fig:read_order}
    \end{figure}

@@

    The main simulated thread in the Memory Manager can flush or evict data from the
    memory cache. The data flushing simulation
    function takes the amount of data to flush as parameter. While
    the amount of data to flush is not reached and there are dirty
    blocks remaining in cache, this function traverses the sorted
    inactive list, then the sorted active list, and writes the
    least recently used dirty block to disk, having set its dirty
    flag to 0. In case the amount of data to flush requires that a
    block be partially flushed, the block is split in two blocks,
    one that is flushed and one that remains dirty. Flushing time
    to disk is simulated by the storage model.

    The cache eviction simulation also runs in
    the main thread. It frees up the page cache by traversing and deleting
    least recently used clean data blocks in the inactive list.
    The amount of data to evict is passed as a parameter and data blocks are deleted
    from the inactive list until the evicted data reaches the required amount,
    or until there is no clean block left in the list.
    If the last evicted block does not have to be entirely evicted, the block is split in two blocks,
    and only one of them is evicted.
    The cache eviction simulation does not contribute to the simulated time
    since cache eviction time is negligible in real systems \tristan{a reference would be welcome}.

@@

    The write simulation described above models page cache in writeback
    mode. Our model also includes a write function in writethrough mode,
    which simply simulates a disk write with the amount of data passed in,
    then evicts cache if needed and adds the written data to the cache.

@@

            Since bandwidth sharing is not simulated, the prototype does not support
            concurrency: it is limited to single-threaded applications running on systems
            with a single-core CPU. We used this prototype for a first validation of our simulation
            model against a real sequential application running on a real system.
            The source code this prototype is available at
            \url{https://github.com/big-data-lab-team/paper-io-simulation/tree/master/exp/pysim}. We used Python 3.7.

            We also implemented our model as part of \wrench, enhancing its
            internal implementation and APIs with a page cache abstraction,
            and allowing users to activate the feature via a command-line
            argument. For the experiments in this paper we use \wrench 1.6
            at commit
            \href{https://github.com/wrench-project/wrench/tree/67185374330d2c4bf274fce222c937e838df5b03}{6718537433},
            which uses \simgrid 3.25, available at
            \url{https://framagit.org/simgrid/simgrid}. Our implementation
            is now part of \wrench's master branch and will be available to
            users with the next \wrench release. W\wrench provides a full \simgrid-based simulation 
            environment that supports, among other features, concurrent accesses to storage devices, 
            applications distributed on multiple hosts, network transfers, 
            and multi-threading.

>>>>

        \subsection{Simulation}

@@

    The Memory Manager simulates two parallel threads: the main one
    implements flushing, eviction, and cached I/Os synchronously, whereas
    the second one, which operates in the background, periodically searches for
    expired dirty data in page cache LRU lists and flushes data to disk. We
    use existing storage simulation models~\cite{lebre2015} to simulate disk and
    memory, characterized by their storage capacity, read and write
    bandwidths, and latency. These models account for
    bandwidth sharing between concurrent memory accesses and concurrent disk accesses.

@@

    In the Linux kernel, page cache LRU lists contain file pages. However,
    due to the large number of file pages, simulating lists of pages
    induces substantial overhead.
    Therefore, we introduce the concept of a data block as a unit to represent data
    cached in memory. A data block is a subset of file pages stored in
    page cache that were accessed in the same I/O operation.
    A data block has information about the file name, block size, last access
    time, a dirty flag that represents whether the data is clean (0)
    or dirty (1), and an entry (creation) time.
    Blocks can have different sizes and a given file can have multiple
    data blocks in page cache. In addition, a data block can be split into an
    arbitrary number of smaller blocks.
    \begin{figure}
           \centering
           \includegraphics[width=\columnwidth]{figures/lru_lists.pdf}
           \caption{Model of page cache LRU lists with data blocks.}
           \label{fig:lrulist}
    \end{figure}

@@

    Our simulation model supports chunk-by-chunk file accesses
    with a user-defined chunk size. However, for simplicity, we assume that file pages are
    accessed in a round-robin fashion rather than fully randomly.
    Therefore, when a file is read, cached data is read only after all uncached data was read, and data from the inactive list is read
    before data from the active list
    (data reads occur from left to right in Figure~\ref{fig:read_order}).
    When a chunk of \emph{uncached} data is read, a new clean block is created
    and appended to the inactive list.
    When a chunk of \emph{cached} data is read, one or more existing data blocks in the LRU lists are accessed.
    If these blocks are clean, we merge them together, update the access time and size of the resulting block,
    and append it to the active list.
    If the blocks are dirty, we move them independently to the active list, to preserve their entry time.
    Because the chunk and block sizes may be different, there are situations
    in a block is not entirely read.
    In this case, the block is split in two smaller blocks and one of them is re-accessed.
    % From Tristan: not sure where to put this nor if it's necessary:
    % , in which chunks are read/written
    % until file is entirely read/written.
    \begin{figure}
           \centering
           \includegraphics[width=\columnwidth]{figures/read_order.pdf}
           \caption{File data read order. Data is read from left to right: uncached data
           is read first, followed by data from the inactive list, and finally data from the active list. }
           \label{fig:read_order}
    \end{figure}

@@

    The main simulated thread in the Memory Manager can flush or evict data from the
    memory cache. The data flushing simulation
    function takes the amount of data to flush as parameter. While
    the amount of data to flush is not reached and there are dirty
    blocks remaining in cache, this function traverses the sorted
    inactive list, then the sorted active list, and writes the
    least recently used dirty block to disk, having set its dirty
    flag to 0. In case the amount of data to flush requires that a
    block be partially flushed, the block is split in two blocks,
    one that is flushed and one that remains dirty. The time needed
    to flush data to disk is simulated by the storage model.

    The cache eviction simulation also runs in
    the main thread. It frees up the page cache by traversing and deleting
    least recently used clean data blocks in the inactive list.
    The amount of data to evict is passed as a parameter and data blocks are deleted
    from the inactive list until the evicted data reaches the required amount,
    or until there is no clean block left in the list.
    If the last evicted block does not have to be entirely evicted, the block is split in two blocks,
    and only one of them is evicted.
    The overhead of the cache eviction algorithm is not part of the simulated time
    since cache eviction time is negligible in real systems \tristan{a reference would be welcome}.

@@

    The above models page cache in writeback
    mode. Our model also includes a write function in writethrough mode,
    which simply simulates a disk write with the amount of data passed in,
    then evicts cache if needed and adds the written data to the cache.

@@

            This prototype does not simulate  bandwidth sharing and thus does not support
            concurrency: it is limited to single-threaded applications running on systems
            with a single-core CPU. We used this prototype for a first validation of our simulation
            model against a real sequential application running on a real system.
            The source code this prototype is available at
            \url{https://github.com/big-data-lab-team/paper-io-simulation/tree/master/exp/pysim}. We used Python 3.7.

            We also implemented our model as part of \wrench, enhancing its
            internal implementation and APIs with a page cache abstraction,
            and allowing users to activate the feature via a command-line
            argument. For the experiments in this paper we use \wrench 1.6
            at commit
            \href{https://github.com/wrench-project/wrench/tree/67185374330d2c4bf274fce222c937e838df5b03}{6718537433},
            which uses \simgrid 3.25, available at
            \url{https://framagit.org/simgrid/simgrid}. Our implementation
            is now part of \wrench's master branch and will be available to
            users with the upcoming \wrench 1.8 release. \wrench provides a full \simgrid-based simulation 
            environment that supports, among other features, concurrent accesses to storage devices, 
            applications distributed on multiple hosts, network transfers, 
            and multi-threading.
